\papertitle

### 2014-03-21 final-report/paper.tex

We propose \system, a visualization tool that helps database developers
explore and understand query execution and data movement in a distributed
database management system (DDBMS). \system provides quick insight into common
problems such as data skew or performance bottleneck by leveraging
visualization techniques to present: (1) data flow between query operators and
between workers, (2) query execution and operator dependencies, (3) cluster
utilization and (4) network utilization.

In particular, \system is built to inspect query execution in Myria~\cite{myria},
a distributed big data management system currently being developed in the
UW CSE database group. Myria aims towards building a distributed database
platform to provide \emph{big data management and analytics as a service} primarily
for scientific applications. \system was designed to be easily ported to
other DDBMS as well (e.g. Spark, Hadoop).

@@

\begin{itemize}
\item \emph{How does the data flow through the workers between query fragments?} 
\item \emph{How can we identify data skew and execution skew?} 
\item \emph{What causes performance bottlenecks?}
\end{itemize}

@@

\textbf{Graph Navigation:} A query plan can be arbitrarily large. Thus we offer two mechanisms that facilitate exploration of the graph for the user: (1) expanding/reducing fragments and (2) zooming. The first mechanism can reduce the size of a graph by a constant factor by collapsing operators that compose a fragment into a single fragment node. The user can click once on a collapsed fragment to expand it, and click once on an expanded fragment to collapse it. Expanding or reducing a node can cause large changes in the graph layout as GraphViz changes the graph layout to minimize overall area. This graph re-shuffling is illustrated in Figure~\ref{graph} where expanding Fragment~2 causes the layout to change. To address these layout changes, we implemented transitions to allow the user to track the fragments as those get reshuffled. The second mechanism that facilitates exploration of the graph for the user is zooming. This feature was implemented in D3 and allows the user to quickly jump from one part of the graph to the next, by zooming in and out of the graph view.

@@

\begin{figure}[ht]
  \includegraphics[width=\columnwidth]{images/operators_chart}
  \caption{The operators chart.}
  \label{fig:operators_chart}
\end{figure}

@@

When the user clicks on a matrix cell, she is able to view the breakdown of the worker pair communication in a
time series chart below the matrix. The user can select multiple such cells to compare different time
series. Figure~\ref{fig:ts} shows an example when the user selected two pairs: (8-2) and (7-6). The x-axis shows
time in seconds and y-axis shows number of tuples exchanged between the source and destination workers.
Just looking at the matrix chart it is evident that there is more data exchanged between the source-destination pair
(8-2) than the pair (7-6). But as the time series comparison shows, there is more communication between (7-6)
than (8-2) for the last 2.5 seconds of the query execution. This points to interesting patterns as to the rate at which
work is executed on different workers during different stages of the execution. Distinct points on the lines are highlighted as
black dots. Hovering the cursor over a dot shows the information for that data point in a tooltip. Meanwhile the cells
in the matrix remain highlighted using the same color used to draw the corresponding time series line. This helps the
user keep track of the mapping between the matrix cell and the corresponding time series line. We also provide
a 'clear selection' button that erases everything from the time series chart upon clicking. Finally we provide the
functionality whereby clicking a row or column label selects every cell in that row or column and displays the
corresponding time series in the time series chart.

@@

\begin{figure}[ht]
  \includegraphics[width=\columnwidth]{images/overview_skew}
  \caption{This \overall shows very irregular usage in fragment 2.}
  \label{fig:overview_skew}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \begin{subfigure}[b]{0.49\columnwidth}
    \includegraphics[width=\columnwidth]{images/skew}
    \caption{Data skew.}
    \label{fig:skew}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\columnwidth}
    \includegraphics[width=\columnwidth]{images/no-skew}
    \caption{No data skew.}
    \label{fig:no-skew}
  \end{subfigure}
  \caption{The network view shows a distinctive pattern when there is data skew. Figure~\ref{fig:skew} shows that most tuples are sent to worker 1 and 5 (origin: row, destination: column).}
  \label{fig:network}
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[width=\columnwidth]{images/execution_skew}
  \caption{This \fragment view shows heavy execution skew on worker~5.}
  \label{fig:exec_skew}
\end{figure}

>>>>

\sloppy
We propose \system, a visualization tool that helps database developers explore
and understand query execution and data movement in a distributed database
management system. \system provides quick insight into common problems
such as data skew or performance bottleneck by leveraging visualization
techniques to present: (1) data flow between query operators and between
workers, (2) query execution and operator dependencies, (3) cluster
utilization, (4) network utilization.

\sloppy
In particular, \system is built to inspect query execution in Myria~\cite{myria},
a distributed big data management system currently being developed in the
UW CSE database group. Myria aims towards building a distributed database
platform to provide \emph{big data management and analytics as a service} primarily
for scientific applications. \system was designed to be easily ported to
other DDBMS as well (e.g. Spark, Hadoop).

@@

\begin{itemize}
\item \emph{How does the data flow through the system between operators?} 
The overall data flow in a DDBMS is represented by a \graph. It is important
for the developer to understand the \graph in order to optimize it. To aid in this
understanding, we provide an interactive visualization of the \graph in \system.

\item \emph{How does the cluster visualization vary for each fragment and operator?} 
In \emph{Myria}, a fragment is composed of multiple sequential database operators running on the same worker. 
A programmer will want to understand the cluster utilization at a given time as well as how much time is being spent
executing an operator. \system provides a \fragment view to aid this understanding. The \fragment view visualizes the utilization of the cluster by showing what operator in the fragment each worker is busy working on.

\item \emph{How is the input data partitioned?}
Data skew is an important problem in a DDBMS, which can result in uneven communication between worker threads
residing on different machines. \system provides a view for \network that all-to-all communication   
between two fragments in the \graph.

@@

\textbf{Graph Navigation:} A query plan can be arbitrarily large. Thus we offer two mechanisms that facilitate exploration of the graph for the user: (1) expanding/reducing fragments and (2) paning. The first mechanism can reduce the size of a graph by a constant factor by collapsing operators that compose a fragment into a single fragment node. The user can click once on a collapsed fragment to expand it, and click once on an expanded fragment to collapse it. Expanding or reducing a node can cause large changes in the graph layout as GraphViz changes the graph layout to minimize overall area. This graph re-shuffling is illustrated in Figure~\ref{graph} where expanding Fragment~2 causes the layout to change. To address these layout changes, we implemented transitions to allow the user to track the fragments as those get reshuffled. The second mechanism that facilitates exploration of the graph for the user is paning. This feature was implemented in D3 and allows the user to navigate a graph when it doesn't fit inside of the browser panel.

@@

\begin{figure}[ht]
  \includegraphics[width=\columnwidth]{images/operators-chart}
  \caption{The operators chart.}
  \label{fig:operators_chart}
\end{figure}

@@

When the user clicks on a matrix cell, she is able to view the breakdown of the worker pair communication in a
time series chart below the matrix. The user can select multiple such cells to compare different time
series. Figure~\ref{fig:ts} shows an example when the user selected two pairs: (8-1) and (7-5). The x-axis shows
time in seconds and y-axis shows number of tuples exchanged between the source and destination workers.
Just looking at the matrix chart it is evident that there is more data exchanged between the source-destination pair
(8-2) than the pair (7-6). But as the time series comparison shows, there is more communication between (7-5)
than (8-1) for the last 2.5 seconds of the query execution. This points to interesting patterns as to the rate at which
work is done at different workers at different stages of the execution. Distinct points on the lines are highlighted as
black dots. Hovering the cursor over a dot shows the information for that data point in a tooltip. Meanwhile the cells
in the matrix remain highlighted using the same color used to draw the corresponding time series line. This helps the
user keep track of the mapping between the matrix cell and the corresponding time series line. We also provide
a 'clear selection' button that erases everything from the time series chart upon clicking. Finally we provide the
functionality whereby clicking a row or column label selects every cell in that row or column and displays the
corresponding time series in the time series chart.

@@

\begin{figure}[ht]
  \includegraphics[width=\columnwidth]{images/overview-skew}
  \caption{This \overall shows very irregular usage in fragment 2.}
  \label{fig:overview_skew}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \begin{subfigure}[b]{0.49\columnwidth}
    \includegraphics[width=\columnwidth]{images/skew.pdf}
    \caption{Data skew.}
    \label{fig:skew}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\columnwidth}
    \includegraphics[width=\columnwidth]{images/no-skew.pdf}
    \caption{No data skew.}
    \label{fig:no-skew}
  \end{subfigure}
  \caption{The network view shows a distinctive pattern when there is data skew. Figure~\ref{fig:skew} shows that most tuples are sent to worker 1 and 5 (origin: row, destination: column).}
  \label{fig:network}
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[width=\columnwidth]{images/execution-skew}
  \caption{This \fragment view shows heavy execution skew on worker~5.}
  \label{fig:exec_skew}
\end{figure}
